---
title: "Something’s Fishy"
author: "Carolyn Chen"
date: "11/12/2018"
output: html_document
---
The goals of this case study are to explore factors (including year, location, bottom depth, bottom temperature, surface temperature) that are related to species abundance (measured by kg/hectare in the variable WTCPUE and #/hectare in the variable NUMCPUE). A particular interest is in exploring whether there are any trends in time, temperature, or other factors in the data. 

Fish:Pacific Herring (Clupea pallasii) 
Load in data files
```{r}
library(data.table)
library(magrittr)
library(dplyr)
library(ggplot2)
library(tidyr)
library("naniar") #
library(shiny)
library(leaflet) #
library(randomForest) #
library(rworldxtra) #
library(rworldmap) #
library(gam) #
library(mgcv)

aisland8320 = fread("aleutianislands1983_2000.csv", fill=TRUE)
aisland1416 = fread("aleutianislands2014_2016.csv", fill=TRUE)
aisland0212 = fread("aleutianislands2002_2012.csv", fill=TRUE)
ebs0216 = fread("beringseaslope2002_2016.csv", fill=TRUE)
eb0508 = fread("easternberingseashelf2005_2008.csv", fill=TRUE)
eb0912 = fread("easternberingseashelf2009_2012.csv", fill=TRUE)
eb1316 = fread("easternberingseashelf2013_2016.csv", fill=TRUE)
eb17 = fread("easternberingseashelf2017.csv", fill=TRUE)
goa9099 = fread("goa1990_1999.csv", fill=TRUE)
goa0105 = fread("goa2001_2005.csv", fill=TRUE)
goa0713 = fread("goa2007_2013.csv", fill=TRUE)
goa1517 = fread("goa2015_2017.csv", fill=TRUE)
nbs8217 = fread("northberingseashelf1982_2017.csv", fill=TRUE)
goa8487 = fread("goa1984_1987.csv", fill=TRUE)

```

Combining Dataframes and Filtering for Pacific Herring. Look at missing values.
```{r}
#Eastern Bering Shelf
herring0508 = eb0508 %>% filter(SCIENTIFIC== "Clupea pallasii")
herring0912 = eb0912 %>% filter(SCIENTIFIC == "Clupea pallasii")
herring1316 = eb1316 %>% filter(SCIENTIFIC == "Clupea pallasii") 
herring17 = eb17 %>% filter(SCIENTIFIC == "Clupea pallasii")  

herringeb = rbind(herring0508, herring0912, herring1316, herring17)
herringeb$LATITUDE=as.numeric(herringeb$LATITUDE)
herringeb$LONGITUDE=as.numeric(herringeb$LONGITUDE)
herringeb$STRATUM=as.numeric(herringeb$STRATUM)
herringeb$YEAR=as.numeric(herringeb$YEAR)
herringeb$WTCPUE=as.numeric(herringeb$WTCPUE)
herringeb$NUMCPUE=as.numeric(herringeb$NUMCPUE)
herringeb$SID=as.numeric(herringeb$SID)
herringeb$BOT_DEPTH=as.numeric(herringeb$BOT_DEPTH)
herringeb$BOT_TEMP=as.numeric(herringeb$BOT_TEMP)
herringeb$SURF_TEMP=as.numeric(herringeb$SURF_TEMP)
herringeb$VESSEL=as.numeric(herringeb$VESSEL)
herringeb$CRUISE=as.numeric(herringeb$CRUISE)
herringeb$HAUL=as.numeric(herringeb$HAUL)


herringeb[herringeb == -9999.0] <- NA
gg_miss_upset(herringeb) 

herringeb[] <- lapply(herringeb, function(x) as.numeric(as.character(x)))
herringeb = herringeb %>% mutate(Location = "Eastern Bering Shelf") %>% as.data.frame()
summary(herringeb) 
#Autitean Islands data
herring8320 = aisland8320 %>% filter(SCIENTIFIC== "Clupea pallasii")
herring1416 = aisland1416 %>% filter(SCIENTIFIC== "Clupea pallasii")
herring1416=select(herring1416,-V18)
herring0212 = aisland0212 %>% filter(SCIENTIFIC== "Clupea pallasii")
##coming in as all characters
herringai = rbind(herring8320, herring1416, herring0212) %>% as.data.frame()

#Look at missing values
herringai[herringai == -9999.0] <-NA
gg_miss_upset(herringai)

herringai[] <- lapply(herringai, function(x) as.numeric(x))
herringai = herringai %>% mutate(Location = "Aleutian Islands") %>% as.data.frame()
summary(herringai) 

#Look at GOA data
herring8487 = goa8487 %>% filter(SCIENTIFIC== "Clupea pallasii") 
herring9099 = goa9099 %>% filter(SCIENTIFIC== "Clupea pallasii") 
herring9099$LATITUDE=as.numeric(herring9099$LATITUDE)
herring9099$LONGITUDE=as.numeric(herring9099$LONGITUDE)
herring9099$STRATUM=as.numeric(herring9099$STRATUM)
herring9099$YEAR=as.numeric(herring9099$YEAR)
herring9099$WTCPUE=as.numeric(herring9099$WTCPUE)
herring9099$NUMCPUE=as.numeric(herring9099$NUMCPUE)
herring9099$SID=as.numeric(herring9099$SID)
herring9099$BOT_DEPTH=as.numeric(herring9099$BOT_DEPTH)
herring9099$BOT_TEMP=as.numeric(herring9099$BOT_TEMP)
herring9099$SURF_TEMP=as.numeric(herring9099$SURF_TEMP)
herring9099$VESSEL=as.numeric(herring9099$VESSEL)
herring9099$CRUISE=as.numeric(herring9099$CRUISE)
herring9099$HAUL=as.numeric(herring9099$HAUL)
herring0105 = goa0105 %>% filter(SCIENTIFIC== "Clupea pallasii")
herring0713 = goa0713 %>% filter(SCIENTIFIC== "Clupea pallasii") 
herring0713$LATITUDE=as.numeric(herring0713$LATITUDE)
herring0713$LONGITUDE=as.numeric(herring0713$LONGITUDE)
herring0713$STRATUM=as.numeric(herring0713$STRATUM)
herring0713$YEAR=as.numeric(herring0713$YEAR)
herring0713$WTCPUE=as.numeric(herring0713$WTCPUE)
herring0713$NUMCPUE=as.numeric(herring0713$NUMCPUE)
herring0713$SID=as.numeric(herring0713$SID)
herring0713$BOT_DEPTH=as.numeric(herring0713$BOT_DEPTH)
herring0713$BOT_TEMP=as.numeric(herring0713$BOT_TEMP)
herring0713$SURF_TEMP=as.numeric(herring0713$SURF_TEMP)
herring0713$VESSEL=as.numeric(herring0713$VESSEL)
herring0713$CRUISE=as.numeric(herring0713$CRUISE)
herring0713$HAUL=as.numeric(herring0713$HAUL)
herring1517 = goa1517 %>% filter(SCIENTIFIC== "Clupea pallasii") 

herringgoa= rbind(herring8487, herring9099, herring0105,herring0713,herring1517) %>% as.data.frame()


#Look at missing values
herringgoa[herringgoa == -9999.0] <- NA
gg_miss_upset(herringgoa)

herringgoa[] <- lapply(herringgoa, function(x) as.numeric(as.character(x)))
herringgoa = herringgoa %>% mutate(Location = "GOA") %>% as.data.frame()
summary(herringgoa)
        
#North Bering Sea
herring8217 = nbs8217 %>% filter(SCIENTIFIC== "Clupea pallasii") %>% mutate(Location = "North Bering Shelf")  %>% as.data.frame()

#Look at missing values
herring8217[herring8217 == -9999.0] <- NA
gg_miss_upset(herring8217)

summary(herring8217)

#East Bering Sea Slope
ebs0216 = ebs0216 %>% filter(SCIENTIFIC== "Clupea pallasii") %>% mutate(Location = "East Bering Sea Slope") %>% as.data.frame()

#Look at missing values
ebs0216[ebs0216 == -9999.0] <- NA
gg_miss_upset(ebs0216)

summary(ebs0216)


```


EDA:
```{r}
#Total Herring from Eastern Bering Shelf, GOA, Aleutian Islands, East Bering Sea Slope, North Bering Shelf
totalherring = rbind(herringeb, herringgoa,herringai,herring8217,ebs0216)
totalherring$Location[totalherring$Location == "Aleutian Islands"] <- 1
totalherring$Location[totalherring$Location == "Eastern Bering Shelf"] <- 2
totalherring$Location[totalherring$Location == "GOA"] <- 3
totalherring$Location[totalherring$Location == "North Bering Shelf"] <- 4
totalherring$Location[totalherring$Location == "East Bering Sea Slope"] <- 5

totalherring=select(totalherring,-SID)
pacificherring=data.frame(totalherring)


totalherring[] <- lapply(totalherring, function(x) as.numeric(as.character(x)))
#Trends in time
#First look at number of entries by year for all herring sites
aggregate(cbind(count = NUMCPUE) ~ YEAR, 
          data = totalherring, 
          FUN = function(x){NROW(x)})


#Look at catch size across years for a given location. 
#First look at example without treating outliers.
p = ggplot(herringeb, aes(x=as.factor(YEAR), y=as.numeric(NUMCPUE))) + geom_boxplot() +ylab("NUMCPUE") +xlab("Year") 
p

#Use outlier method to remove extreme values. We will assume these are due to improper imputation.
#NUMCPUE with Tukey's Outliers removed
remove_outliers1 <- function(x) {
lowerq = quantile(as.numeric(x$NUMCPUE),na.rm=TRUE)[2]
upperq = quantile(as.numeric(x$NUMCPUE),na.rm=TRUE)[4]
iqr = as.numeric(upperq - lowerq)
extreme.threshold.upper = (iqr * 3) + upperq
extreme.threshold.lower = lowerq - (iqr * 3)
upbound = x[x$NUMCPUE < as.numeric(extreme.threshold.upper),]
outlierrem = upbound[upbound$NUMCPUE > as.numeric(extreme.threshold.lower),]
p1 = ggplot(outlierrem, aes(x=as.factor(YEAR), y=as.numeric(NUMCPUE))) + geom_boxplot() +ylab("NUMCPUE") +xlab("Year") 
p1
}

remove_outliers1(herringeb)
remove_outliers1(herringgoa)
remove_outliers1(herringai)
remove_outliers1(herring8217)
remove_outliers1(ebs0216)

#WTCPUE graph with Tukey's outliers removed
remove_outliers2 <- function(x) {
lowerq = quantile(as.numeric(x$WTCPUE),na.rm=TRUE)[2]
upperq = quantile(as.numeric(x$WTCPUE),na.rm=TRUE)[4]
iqr = as.numeric(upperq - lowerq)
extreme.threshold.upper = (iqr * 3) + upperq
extreme.threshold.lower = lowerq - (iqr * 3)
upbound = x[x$WTCPUE < as.numeric(extreme.threshold.upper),]
outlierrem = upbound[upbound$WTCPUE > as.numeric(extreme.threshold.lower),]
p1 = ggplot(outlierrem, aes(x=as.factor(YEAR), y=as.numeric(WTCPUE))) + geom_boxplot() +ylab("WTCPUE") +xlab("Year") 
p1
}

remove_outliers2(herringeb)
remove_outliers2(herringgoa)
remove_outliers2(herringai)
remove_outliers2(herring8217)
remove_outliers2(ebs0216)

remove_outliers3 <- function(x) {
lowerq = quantile(as.numeric(x$NUMCPUE),na.rm=TRUE)[2]
upperq = quantile(as.numeric(x$NUMCPUE),na.rm=TRUE)[4]
iqr = as.numeric(upperq - lowerq)
extreme.threshold.upper = (iqr * 3) + upperq
extreme.threshold.lower = lowerq - (iqr * 3)
upbound = x[x$NUMCPUE < as.numeric(extreme.threshold.upper),]
outlierrem = upbound[upbound$NUMCPUE > as.numeric(extreme.threshold.lower),]
return(outlierrem)
}

herringebR = remove_outliers3(herringeb)
herringgoaR = remove_outliers3(herringgoa)
herringaiR = remove_outliers3(herringai)
herringnbsR = remove_outliers3(herring8217)
herringebsR = remove_outliers3(ebs0216)

totalherringR = rbind(herringebR,herringgoaR,herringaiR,herringnbsR,herringebsR)
#Bottom Temperature
p4= ggplot(herringeb, aes(x=as.factor(YEAR), y=as.numeric(BOT_TEMP))) + geom_boxplot() +ylab("Bottom Temperature") +xlab("Year") + ggtitle("Bottom Temperature of East Bering Shelf")
p4 
p51= ggplot(herring8217, aes(x=as.factor(YEAR), y=as.numeric(BOT_TEMP))) + geom_boxplot() +ylab("Bottom Temperature") +xlab("Year") + ggtitle("Bottom Temperature of North Bering Sea")
p51 
p52 = ggplot(herringgoa, aes(x=as.factor(YEAR), y=as.numeric(BOT_TEMP))) + geom_boxplot() +ylab("Bottom Temperature") +xlab("Year") + ggtitle("Bottom Temperature of GOA")
p52
p53 = ggplot(herringai, aes(x=as.factor(YEAR), y=as.numeric(BOT_TEMP))) + geom_boxplot() +ylab("Bottom Temperature") +xlab("Year") + ggtitle("Bottom Temperature of Aleutian Islands")
p53
p54 = ggplot(ebs0216, aes(x=as.factor(YEAR), y=as.numeric(BOT_TEMP))) + geom_boxplot() +ylab("Bottom Temperature") +xlab("Year") + ggtitle("Bottom Temperature of East Bering Sea Slope")
p54


#Location of fish
newmap <- getMap(resolution = "high")
plot(newmap)
points(totalherring$LONGITUDE, totalherring$LATITUDE, col = "blue", cex = .8)

```


Model for Species Abundance
```{r, warnings=FALSE}

#Generalized Additive Model
gam1<-gam(NUMCPUE~LATITUDE + LONGITUDE + s(STRATUM) + s(YEAR) + s(BOT_DEPTH) + s(BOT_TEMP) + s(SURF_TEMP) + Location ,data = totalherringR)
summary(gam1)

gam1<-gam(WTCPUE~LATITUDE + LONGITUDE + s(STRATUM) + s(YEAR) + s(BOT_DEPTH) + s(BOT_TEMP) + s(SURF_TEMP) + Location ,data = totalherringR)
summary(gam1)

#Generalized Linear Models
#First look at normal model
fit <- glm(NUMCPUE~LATITUDE + LONGITUDE + STRATUM + YEAR + BOT_DEPTH + BOT_TEMP + SURF_TEMP + Location, data=totalherringR, family=gaussian(link="identity"))
summary(fit)

fit <- glm(WTCPUE~LATITUDE + LONGITUDE + STRATUM + YEAR + BOT_DEPTH + BOT_TEMP + SURF_TEMP + Location, data=totalherringR, family=gaussian(link="identity"))
summary(fit)

#Then look at a Gamma model
fit <- glm(NUMCPUE~LATITUDE + LONGITUDE + STRATUM + YEAR + BOT_DEPTH + BOT_TEMP + SURF_TEMP + Location, data=totalherringR, family=Gamma(link="inverse"))
summary(fit)

fit <- glm(WTCPUE~LATITUDE + LONGITUDE + STRATUM + YEAR + BOT_DEPTH + BOT_TEMP + SURF_TEMP + Location, data=totalherringR, family=Gamma(link="inverse"))
summary(fit)


```