---
title: "Getting The Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How to Obtain Data Frames From CIA Factbook
First, you will need to obtain the factbook from https://www.cia.gov/library/publications/download/. Click "factbook.zip" and this will download the entire factbook for you. (Note: this code corresponds to the 2014 factbook, but similar principles should be applicable to other factbooks). Then you have two options for obtains your data sets:

## Option 1: Rank Order Txt Files
Your first option is using the rank order section of the factbook which has txt files for each variable. However, the significant drawnback of this is that it only has one value for each country, i.e. you cannot use it to obtain a variable such as population structure since it has multiple values for each country. 
`
- Remember to give your columns names.
- To find the corresponding number for each variable, e.g. 2119 for population, you use the "rankorderguide.html" and click on your desired variable. Then in the address bar at the top of the screen you can see the correct number for it.
- You may need to delete columns/clean the data to obtain the correct format.

Note: You will need to iterate this process for all desired variables & merge the data sets to replicate the full.RData file. 
 
```{r}
library(tidyverse) # We will be importing our data using the tidyverse library

# get country codes
ctr_codes <- read.csv("country_codes.csv", header = TRUE)
cia_codes <- ctr_codes[,1:2]

# population Data
popDF <- read_delim("factbook/rankorder/rawdata_2119.txt", delim = "\t", col_names = c("X1", "country","pop"), col_types = list(col_integer(), col_character(), col_number()))
# Hint: you may want to also explore the read_table function for your homework
popDF <- popDF[,-1] # remove blank column
popDF <- merge(cia_codes, popDF, by.x="Country", by.y = "country") 
popDF <- popDF[,2:3]
names(popDF) <- c("ctry", "pop")
save(popDF, file="popDF.RData")

# Infant Mortality Data 
infMortDF <- read_delim("factbook/rankorder/rawdata_2091.txt", delim = "\t", col_names = c("X1", "country","infMort"), col_types = list(col_integer(), col_character(), col_number()))
infMortDF <- infMortDF[,-1]
infMortDF <- merge(cia_codes, infMortDF, by.x="Country", by.y = "country")
infMortDF <- infMortDF[,2:3]
names(infMortDF) <- c("ctry", "infMort")
save(infMortDF, file="infMortDF.RData")

```
## Option 2: Field HTML Files
Your second option is to use the field html files. These have the advantage of being more complete and more detailed information on each variable. However, they are also in a less user-friendly format.

```{r}
library(rvest)

# read in data
test <- read_html("factbook/fields/print_2010.html")
t <- test %>% html_table(fill = TRUE) # converts html to a list

# Extracting the desired data
# First view the data and determine what parts of the data we need
# head(t)

# create an empty data frame to hold our data
population_structure <- matrix(0, 230, 6)
colnames(population_structure) <- c("country", "ps_0-14", "ps_15-24", "ps_25-54", "ps_55-64", "ps_65+")

# filling the data frame
for(i in 4:233){
  # getting country
  population_structure[i-3, 1] <- t[[i]][["X1"]][2]
  # splitting based on category
  levels <- strsplit(t[[i]][["X2"]][2], "\n")
  for(j in 1:5){
    # obtaining the relevant sections of the data
    start <- regexpr(":",levels[[1]][j]) + 2
    end <- regexpr("%",levels[[1]][j]) - 1
    population_structure[i-3, j+1] <- as.numeric(substr(levels[[1]][j], start, end))
  }
}

# Another Example
test <- read_html("factbook/fields/print_2091.html")
t <- test %>% html_table(fill = TRUE)
IM <- matrix(0, 233, 4)
colnames(IM) <- c("country", "IM_total", "IM_male", "IM_female")
for(i in 4:236){
  IM[i-3, 1] <- t[[i]][["X1"]][2]
  levels <- strsplit(t[[i]][["X2"]][2], "\n")
  for(j in 1:3){
    start <- regexpr(":",levels[[1]][j]) + 2
    end <- regexpr("d",levels[[1]][j]) - 2
    IM[i-3, j+1] <- as.numeric(substr(levels[[1]][j], start[1], end[1]))
  }
}
```

 