---
title: Cox Model and Missing Data
output: 
  revealjs::revealjs_presentation:
    theme: night
    highlight: espresso
    center: true
    transition: none
    css: styles.css
    fig_caption: true
    reveal_options:
      progress: true
      slideNumber: true
      
  
---


## Ovarian data

Recall the ovarian data: we will create a new dataset just with our redefined predictors for simplicity.

```{r ovarian2}
data(ovarian); library(mice); library(survival)
ovarian$residyes=ovarian$resid.ds-1 #2=yes goes to 1 and 1=no to 0
ovarian$trt1=2-ovarian$rx #1 stays 1 and 2 goes to 0
ovarian$ecog1=2-ovarian$ecog.ps
ovarian$age10=ovarian$age/10 #10-year difference for better interpretability
ovarian2=ovarian[,c(1:2,7:10)] #pick off just vars of interest
```


## Missing data in survival analysis

Suppose some of the variables in the ovarian data were subject to missing values. For example, perhaps the residual disease indicator has some missing values generated at random.

```{r genmissing}
set.seed=1213  #for reproducibility
#randomly sample 5 indices to set to missing (MCAR)
msng=round(runif(5,min=1,max=length(ovarian$residyes)))
#set residyes values corresponding to those indices to missing
ovarian2$residyes[msng]=NA
```

## Look at data

See how many NA values we have....

```{r seedata}
table(ovarian2$residyes,useNA="always")
```

## Just for comparison: complete case analysis

```{r ccase}
coxph(Surv(futime,fustat)~age10+residyes+trt1+ecog1,data=ovarian2)
```

## Multiple imputation

```{r mi}
imp=mice(ovarian2,m=20)
coximpute=with(imp,coxph(Surv(futime,fustat)~age10+residyes+trt1+ecog1))
```

## Results
```{r pooling}
#note the pool function assumes infinite error df; not appropriate in small samples
#(this is a small sample!)
summary(pool(coximpute))
```

You can see only age is statistically significant using $p<0.05$ as a criterion in this analysis.

